---
title: "Nonlinear"
author: "Yang Wang"
date: "2/20/2023"
output: pdf_document
---

```{r}
# illustrate the use of nonlieaner models in R
```


```{r}
library(ISLR)
data("Wage")
```


# Polynomial Regression
The poly() function generates a basis of orthoganal polynomials.
```{r}
fit = lm(wage~poly(age,4), data = Wage)
summary(fit)
```

let's make a plot of the fitted function, along with the standard errors of the fit.
```{r, fig.width = 7, fig.height = 5}
agelims = range(Wage$age)
age.grid = seq(from = agelims[1], to=agelims[2])
preds = predict(fit, newdata = list(age = age.grid), se = TRUE)
se.band = cbind(preds$fit + 2*preds$se.fit, preds$fit - 2*preds$se.fit)
plot(Wage$age, Wage$wage, col = "darkgrey")
lines(age.grid, preds$fit, lwd = 2, col = "blue")
matlines(age.grid, se.band, lty = 2, col = "blue", lwd = 2)
```

Other more direct ways of doing this in R.
"age^2" means something else to the formula language. 
I() is a "wrapper" function. 
I(age^2) is protected. 

```{r}
fita = lm(wage~age+I(age^2)+I(age^3)+I(age^4), data = Wage)
summary(fita)
```

 two method provide the same fits
```{r}
plot(fitted(fit), fitted(fita))
```

compare models using the function anova()
```{r}
fita = lm(wage ~ education, data = Wage)
fitb = lm(wage ~ education + age, data = Wage)
fitc = lm(wage ~ education + poly(age,2), data = Wage)
fitd = lm(wage ~ education + poly(age,3), data = Wage)
anova(fita, fitb, fitc, fitd)

```

# Polynomial logistic regression
Now we fit a logistic regression model to a binary response variable.
we code the big earners (>250k) as 1, else 0

```{r}
fit = glm(I(wage>250)~ poly(age,3), data = Wage, family = binomial)
summary(fit)
preds = predict(fit, list(age = age.grid), se = T)
se.bands = preds$fit + cbind(fit = 0,lower = -2*preds$se.fit, upper = 2*preds$se)
```


We have done the computation (fit, confidence interval) on the logit scale. To transform we need to apply the inverse logit mapping
$p = \frac{e^\eta}{1+e^\eta}$
```{r}
prob.bands = (exp(se.bands))/(1+exp(se.bands))
matplot(age.grid, prob.bands, col = "blue", lwd = c(2,1,1), lty = c(1,2,2), type = "l", ylim = c(0,0.1))
points(jitter(Wage$age), I(Wage$wage>250)/10, pch = "|", cex = 0.5)
```


# Splines

Splines are more flexible than polynomials, but the idea is rather similar.

```{r}
library(splines)
fit = lm(wage ~ bs(age, df = 3, knots = c(25,40,60)), data = Wage)
plot(Wage$age, Wage$wage, col = "darkgrey")
lines(age.grid, predict(fit, list(age = age.grid)), col = "blue")
abline(v=c(25,40,60), lty = 2, col = "blue")
```
# Natural Spline
```{r}
library(splines)
fit = lm(wage ~ ns(age, df = 3, knots = c(25,40,60)), data = Wage)
plot(Wage$age, Wage$wage, col = "darkgrey")
lines(age.grid, predict(fit, list(age = age.grid)), col = "blue")
abline(v=c(25,40,60), lty = 2, col = "blue")
```
# smoothing spline
It doen't require knot selection, but it dose have a smoothing parameter, which can conviniently be secified via the effective degrees of freedom or "df"

```{r}

plot(Wage$age, Wage$wage, col = "darkgrey")
lines(age.grid, predict(fit, list(age = age.grid)), col = "blue")
abline(v=c(25,40,60), lty = 2, col = "blue")

fit = smooth.spline(Wage$age, Wage$wage, df = 16)
lines(fit,col = "red", lwd = 2)
```

Or we can use LOO cross-validation to select the smoothing parameter for us automatically

```{r}
fit = smooth.spline(Wage$age, Wage$wage, cv = TRUE)
fit
```

# Generalized Additive Model
So for we have focused on fitting models with mostly single nonlinear terms.
The "gam" package make it easier to work with multiple nonlinear terms. In addition, it knows how to plot these functions and their standard errors.

```{r}
library(gam)
gam1 = gam(wage ~ s(age, df =4) + s(year, df = 4)+education, data = Wage)
par(mfrow = c(1,3))
plot(gam1, se = T)
gam2 = gam(I(wage>250) ~ s(age, df =4) + s(year, df = 4)+education, data = Wage)
plot(gam2, se = T)
```

Let's see if we need a nonlinear term for year

```{r}
gam2a = gam(I(wage>250) ~ s(age, df =4) + year +education, data = Wage)
anova(gam2, gam2a)
AIC(gam2)
AIC(gam2a)
```

One nice feature if the "gam" package is that it knows how to plot the function nicely, even for modes fit by 'lm' and "glm'

```{r}
par(mfrow = c(1,3))
lm1 = lm(wage ~ ns(age, df = 4)+ ns(year, df = 4) + education, data = Wage)
plot.Gam(lm1)

```